---
title: "Katahdin power analysis"
format: pdf
---

## Relationship between strata and climate variables

We will consider two climate variables: temperature (`tmp`) and relative humidity (`rh`). There are 16 strata and we will investigate two different scenarios:

1. `tmp` and `rh` have strong correlation to strata 
1. `tmp` and `rh` have weaker correlation to strata 

The scenario of no corelation is unlikely given that strata were choosen to represent meaningful climatic variation.


```{r}
#| echo: false
#| message: false

# general set-up
library(ggplot2)
library(VGAM)
library(cowplot)

# helper function to generate climate data from strata
clm_strata <- function(b_tmp, b_rh, sd, n) {
    s <- 1:16
    srep <- rep(s, each = n)
    tmp <- rnorm(length(srep), b_tmp * srep, sd)
    rh <- rnorm(length(srep), b_rh * srep, sd)
    
    data.frame(s = srep, tmp = scale(tmp), rh = scale(rh))
}



d1 <- clm_strata(b_tmp = 2, b_rh = -2, sd = 2, n = 4)
d2 <- clm_strata(b_tmp = 0.5, b_rh = -0.5, sd = 2, n = 4)

d1$scenario <- "strong response"
d2$scenario <- "weaker response"
d <- rbind(d1, d2)

p_tmp <- ggplot(d, aes(x = s, y = tmp, color = s)) +
    geom_jitter(height = 0) +
    facet_wrap(vars(scenario), nrow = 1) +
    scale_color_viridis_c() +
    theme(legend.position = "none") +
    xlab("Strata") + ylab("Scaled temperature")

p_rh <- ggplot(d, aes(x = s, y = rh, color = s)) +
    geom_jitter(height = 0) +
    facet_wrap(vars(scenario), nrow = 1) +
    scale_color_viridis_c() +
    theme(legend.position = "none") +
    xlab("Strata") + ylab("Scaled relative humidity")

plot_grid(p_tmp, p_rh, nrow = 2)
```



## Setting up possible scenarios for the relationships between focal taxa and climate

We will assume there is an optimal climate for the taxon where its percent cover peaks as displayed in the below figure which shows an idealized response of percent cover to climate (i.e. without random noise).


```{r}
pfun <- function(x, b0, b1, b2) {
    1 / (1 + exp(-(b0 + b1 * x + b2 * x^2)))
}

x <- d1$tmp

ideal_response <- data.frame(s = d1$s, x = x, p = pfun(x, 2, 0.5, -4))

ggplot(ideal_response, aes(x = x, y = p * 100, color = s)) +
    geom_point() +
    scale_color_viridis_c(name = "Strata") +
    xlab("Scaled temperature or relative humidity") +
    ylab("Percent cover")
```

Different taxa will have different optimal climates, different maximum percent covers, and different tolerances for climatic conditions away from their optimal climate (i.e. narrow or wider peaks in percent cover).

The goal of our power analysis is to determine what sample size is necessary to accurately capture the climatic preferences of focal taxa under different representative scenarios of patchy versus ubiquitous spatial distributions and climate niches that are narrow versus wide.

We will model the patchiness of taxa via a zero inflated binomial distribution. The below figure shows the difference between a patchy spatial distribution and a ubiquitous spatial distribution. 


```{r}
#| echo: false

# general simulation set-up ----

# function to make zero inflated binomial data
zbdata <- function(x1, x2, r0, size, a0, a1, a2, b1, b2) {
    nn <- length(x1)
    
    zdata <- data.frame(pstr0 = logitlink(r0, inverse = TRUE),
                        x1 = x1,
                        x2 = x2,
                        sv = rep(size, length = nn))
    
    zdata$mubin <- logitlink(a0 + a1 * x1 - a2 * x1^2 + 
                                 b1 * x2 - b2 * x2^2, 
                             inverse = TRUE)
    
    zdata$y <- rzibinom(nn, 
                        size = zdata$sv, 
                        prob = zdata$mubin, 
                        pstr0 = zdata$pstr0)
    
    return(zdata)
}

# data.frame of params for different scenarios 
pars <- matrix(c(-2, 2, 0.5, 4, 0.5, 4, 
                 1, 1, 0.5, 4, 0.5, 4, 
                 -2, 0.5, 0.05, 0.1, 0.05, 0.1, 
                 1, 0.25, 0.05, 0.1, 0.05, 0.1), 
               byrow = TRUE, nrow = 4) |> 
    as.data.frame()

names(pars) <- c("r0", "a0", "a1", "a2", "b1", "b2")

pars <- cbind(abund_scenario = c("splst_u", 
                                 "splst_r", 
                                 "gnlst_u", 
                                 "gnlst_r"), 
              pars)



# simulate data under all scenarios
sim_dat <- function(d1, d2) {
    o <- lapply(1:nrow(pars), function(i) {
        o1 <- zbdata(x1 = d1$tmp, x2 = d1$rh, 
                     r0 = pars[i, "r0"], size = 100, 
                     a0 = pars[i, "a0"], a1 = pars[i, "a1"], a2 = pars[i, "a2"], 
                     b1 = pars[i, "b1"], b2 = pars[i, "b2"])
        
        o2 <- zbdata(x1 = d2$tmp, x2 = d2$rh, 
                     r0 = pars[i, "r0"], size = 100, 
                     a0 = pars[i, "a0"], a1 = pars[i, "a1"], a2 = pars[i, "a2"], 
                     b1 = pars[i, "b1"], b2 = pars[i, "b2"])
        
        o1$scenario <- paste("senv", pars[i, "abund_scenario"], sep = "_")
        o2$scenario <- paste("wenv", pars[i, "abund_scenario"], sep = "_")
        
        return(rbind(o1, o2))
    })
    
    return(do.call(rbind, o))
}

set.seed(20)
all_dat <- sim_dat(d1, d2)

# example plot ----

lab_map <- c(senv_splst_r = "Patchy",
             senv_splst_u = "Ubiquitous")

ggplot(all_dat[all_dat$scenario %in% c("senv_splst_u", "senv_splst_r"), ], 
             aes(x = x1, y = y)) +
    facet_grid(cols = vars(scenario), 
               labeller = labeller(scenario = lab_map)) + 
    geom_point() +
    xlab("Scaled temperature or relative humidity") +
    ylab("Percent cover")
```

What we can see is that the patchy taxon still peaks at the same climate optimum, but there are more sites where it was absent due to its patchy distribution. The ubiquitous taxon has fewer sites where it is absent, but there are still some gaps because we assume all taxa have some about of patchiness.


## Power analysis for relationship between focal taxa and climate

Now we investigate our statistical power to detect a response of the percent cover of a focal taxon to local climate variation, and how that power changes with number of sampling sites.  To recap, we are investigating 8 scenarios coming from all unique combinations of the following conditions:

- *Variation in climate across strata*: 
    1. temperature and relative humidity have strong correlation with strata 
    1. temperature and relative humidity have weaker correlation with strata 
- *Climate niche*:
    1. narrow 
    2. wide
- *Spatial distribution*:
    1. patchy
    2. ubiquitous 

The below figure shows how statistical power increases with the number of sample replicates per stratum under all combinations of those conditions.

```{r}
#| echo: false
#| message: false
#| cache: true


# function to generate a simulated full data set including climate-strata 
# relationship and percent covers, fit model, compute LRT p-val
#' @param n is number of replicates per stratum
#' @param ncore is number of cores to use in multithreaded simulation
#' @param B is number of simulation replicates for calculating power

full_sim <- function(n, ncore = 8, B = 500) {
    # o <- lapply(1:B, function(i) {
    o <- parallel::mclapply(1:B, mc.cores = ncore, function(i) {
        d1 <- clm_strata(b_tmp = 2, b_rh = -2, sd = 2, n = n)
        d2 <- clm_strata(b_tmp = 0.5, b_rh = -0.5, sd = 2, n = n)
        
        dat <- sim_dat(d1, d2)
        
        pvals <- sapply(unique(dat$scenario), function(s) {
            this_dat <- dat[dat$scenario == s, ]
            
            head(this_dat)
            
            fit <- try(vglm(cbind(y, sv - y) ~ x1 + I(x1^2) + x2 + I(x2^2), 
                            family = zibinomialff, data = this_dat))
            
            fit0 <- try(vglm(cbind(y, sv - y) ~ 1, 
                             family = zibinomialff, data = this_dat))
            
            if("try-error" %in% class(fit) | "try-error" %in% class(fit0)) {
                return(NA)
            } else {
                lrt <- -2 * (logLik(fit0) - logLik(fit))
                
                return(pchisq(lrt, df = 4, lower.tail = FALSE))
            }
        })
        
        return(pvals < 0.05)
    })
    
    o <- do.call(rbind, o)
    
    # this returns proportion of times we reject null
    return(colMeans(o, na.rm = TRUE))
    
}

# now we can simulate across different sample sizes
nn <- 1:6

pwr <- lapply(nn, full_sim, B = 100)

# data.frame for plotting
final_pwr <- data.frame(n = rep(nn, each = length(pwr[[1]])), 
                        scenario = names(unlist(pwr)), 
                        power = unlist(pwr))
conditions <- strsplit(final_pwr$scenario, split = "_")
conditions <- do.call(rbind, conditions) |> as.data.frame()
names(conditions) <- c("env", "niche", "spatial")

final_pwr <- cbind(final_pwr, conditions)

# plotting 
lab_map <- c(r = "Patchy",
             u = "Ubiquitous", 
             senv = "Strong env corr", 
             wenv = "Weaker env corr")

ggplot(final_pwr, aes(x = n, y = power, color = niche)) +
    geom_hline(yintercept = 0.8) + 
    geom_point() +
    geom_line() + 
    xlab("Sample size per stratum") +
    ylab("Statistical power") +
    scale_color_manual(labels = c("Wide", "Narrow"), 
                       values = scales::hue_pal()(2)) +
    facet_grid(rows = vars(env), cols = vars(spatial), 
               labeller = labeller(env = lab_map[2:4], 
                                   spatial = lab_map[1:2])) +
    theme(panel.grid.major.y = element_blank(), 
          panel.grid.minor.y = element_blank())
```

The horizonal black line shows the 0.8 power cuttoff.  This is a commonly used cuttoff for determining sufficient sample size.  What we see is that it takes a minimum of 4 replicates under all combinations of conditions to reach 0.8 power.  4 replicates per stratum leads precisely to our proposed sample size of `r 4 * 16`